Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-16 extended_catalog/paper/extended_catalog.tex
lots of keiths changes

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\renewcommand{\sec}{\text{s}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}
\newcommand{\loglikelihood}{\ensuremath{LL}\xspace}
\newcommand{\chandra}{\text{{\em Chandra}}\xspace}
\newcommand{\swiftxrt}{\text{{\em Swift}/XRT}\xspace}
\newcommand{\rosat}{\text{{\em ROSAT}}\xspace}
\newcommand{\suzaku}{\text{{\em Suzaku}}\xspace}
\newcommand{\asca}{\text{{\em ASCA}}\xspace}
\newcommand{\einstein}{\text{{\em Einstein}}\xspace}
\newcommand{\xmmnewton}{\text{{\em XMM-Newton}}\xspace}

@@

\begin{abstract}
We present a new method for analyzing 
the spatial extension of sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em Fermi Gamma-ray Space Telescope (Fermi)}. 
The source extension is an important input to correctly 
associate LAT sources with their counterparts at other
wavelengths. We provide
a series of Monte Carlo studies to validate this tool and calculate
the LAT's detection threshold to spatially extended sources.  We then
apply this tool to test all sources in the two year source catalog for
extension. We present on the detection of nine spatially extended sources
in addition to the twelve spatially extended sources reported in the
second Fermi-LAT Catalog (2FGL).
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the Fermi
Gamma-Ray Space Telescope (FGST). It has been surveying the $\gamma$-ray
sky since June 2008.  Fermi has a broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\approx 2.4 \sr$), and large effective area
($\approx 8000 \cm^2$ at $>1 \gev$).

Using two years of all sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources \citep{second_cat}. Many of these can be
positionally associated with a variety of source classes including Active
Galactic Nuclei (AGN), pulsars, galaxies, and supernova remnants (SNRs).
While many of the LAT sources can be spatially resolved when observed
at other frequencies, detection of an extension at \gev energies by
the LAT is complicated by the rather large point-spread function (PSF)
of the instrument. The angular resolution for events converting in the
front part of the detector, defined as the 68\% containment radius for
a point source, is approximately $3.5^{\circ}$ at 100 \mev, improving
to about 0.1$^{\circ}$ at 10 \gev.

However, a number of sources positionally coincident with LAT sources
exhibit extension at other wavelengths which are larger or comparable to
the LAT PSF.  In particular certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites can therefore be expected to be observed as spatially extended
sources in the LAT.  The LAT collaboration has previously reported on the
discovered of five spatially extended supernova remnants, IC443, W28, W44,
W51C, and RX J1713.7-3946 \citep{ic443,w28,w44,w51c,rx_j1713_lat}. In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 \citep{msh1552,velax,fermi_hess_j1825}, as well
as two close-by galaxies, the Large Magellanic and Small Magalenic
Cloud \citep{lmc,smc}, and finally the radio galaxy Centarus A
\citep{cen_a_lat}.

For each of these sources, the extension was determined based on
dedicated analysis of the particular source or region of interest
instead of being the outcome of a systematic search of the LAT catalog.
For the systematic scan of all LAT catalog sources presented here, we
therefore expect to find additional spatially extended LAT sources, in
particular since many extended sources in the Galactic plane have been
detected at \tev energies using Air Cherenkov detectors and \tev and
\gev emission often originates from the same population of high-energy
particles.  A variety of spatially extended sources have been detected
in the Galactic plane at \tev energies. Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which presented 14 spatially extended sources
 with extensions varying from
$~0.1\deg$ to $~0.25\deg$ \citep{hess_plane_survey}.
In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes}
are amongst the few Galactic sources
for which no significant extension was detected at \tev energies. Given
these findings, it can be expected that if the source population at \gev
energies has similar properties, that at least some of the Fermi-LAT
sources in our Galaxy should be spatially resolvable at \gev energies.

@@

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large PSF and the large
candidate source density in our Galaxy, source confusion is a severe
problem for the identification of LAT sources.  Finding a coherent
source extension across different energy bands allows a firm association
of a LAT source to an otherwise confused counterpart.  Because of the
strongly varying PSF of the LAT, the spatial and spectral information
about a source do not simply decouple. A biased spatial model will bias
the spectral model of the source. In particular, performing a spectral
analysis on a spatially extended source modeled as point-like will
systematically shift the spectral index to be softer.  Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Cloud \citep{first_cat}.
Additionally, such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources in the \gev energy range using the
LAT are challenging because of the strongly energy-dependent PSF.
Additional complications arise for sources along the Galactic plane due to
systematic uncertainties in the Galactic diffuse emission.  The LAT's
PSF is limited at lower energies by multiple scattering in the tracker
and the 68\% containment radius of the PSF is approximately 5.1\deg
at 100 \mev  when averaged over instrument acceptance and including
photons which convert in either the thick or thin layers of the
tracker. The PSF improves with energy and approaches a limit given
by the granularity of the silicon strips and is 0.14\deg at 10 \gev
\citep{on_orbit_calibration}.  However, since most high energy
astrophysical sources have power-law spectra, the better resolution of the higher
energy photons is offset by their low
statistics. Therefore a detailed investigation has been performed in this
work to determine the optimal energy range to measure source extension
(see below).

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool
provides a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is maximized,
given a parametrized spatial and spectral model of 
the source and its surrounding region.
The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position,
extension, and spectrum can be simultaneously fit.

This analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center \citep{fssc}} because \gtlike can only 
fit the spectral parameters of the model.

We note that \gtlike has been used in the past in several studies of
source extension in the LAT collaboration \citep{lmc,smc,w28,w51c}.
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort can be
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT collaboration and called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

The choice to base the spatial extension fitting on \pointlike
rather than \gtlike was made on considerations of computing time.
The \pointlike algorithm was optimized for speed to handle
larger numbers of sources efficiently which is important for
our catalog scan as well as the Monte Carlo validation of our method.
Details on the \pointlike package can be found in \citep{matthew_kerr_thesis}.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source \citep{minuit_documentation}.  As will be described in
section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is necessary to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated
from \tsext, which 
is the test statistic of the likelihood ratio test of
a model with a spatially extended source and a model
with a point-like source and defined as
\begin{equation}
  \tsext=2\log(\loglikelihood_\text{ext}/\loglikelihood_\text{ps}).
\end{equation}
where \loglikelihood is the log of the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
section~\ref{monte_carlo_validation}.

For extended sources with an assumed radially symmetric shape,
we use a semi-analytic calculation to speed up the calculation of
the expected photon distribution. The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r'^2 dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a {\em King function}
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be done analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r'^2 dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

There will always be a small numerical discrepancy between the expected
photon distribution derived from a true point source and a very small
extended source due to numerical error in the convolution.  In most
situations, this error would be insignificant.  But in particular for
very bright sources, this numerical error has the potential to bias the
test statistic for the extension test. The source with an extended shape
to the likelihood fitting it with its extension fixed to ${10^{-10}}\deg$.

We estimate the errors on the extension
by recalculating the likelihood for a set of extension values
while keeping the position fixed.

We estimate the error on the extension of a source by fixing the position
of the source and varying the extension until the likelihood has fallen
by $\tfrac{1}{2}$, corresponding to a $1\sigma$ error.  This method is
shown schematically in figure~\ref{four_plots_ic443}(a) which shows the
change in \loglikelihood when varying the extension of the SNR IC443.
The localization error is calculated by fixing the extension and spectrum
of the source and fitting a Gaussian to the likelihood as a function
of position.

@@

\pointlike is important for LAT analysis that require many iterations
such as source localization and extension fitting.  On the other hand,
because of the fewer approximations used in calculating the likelihood
we expect the spectral parameters found with \gtlike to be slightly
more precise.  Furthermore, because \gtlike is the standard likelihood
analysis package, it has been more extensively validated for spectral
analysis.  For that reason, in the following analysis  we use \pointlike
to determinate the position and extension of a source and then determine
the spectrum using \gtlike. Both \gtlike and \pointlike can be 
used to estimate
the statistical significance of the extension of a source and we require
that both methods agree to claim a detection.

We found good agreement between the two methods.  
Unless explicitly mentioned, all \ts, \tsext, and spectral
parameters were calculated using \gtlike using \pointlike's positions
and extensions.

@@

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Like extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a two dimensional Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, the convolution
of them with the PSF look similar.  Figure~\ref{compare_disk_gauss}
demonstrates this by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and the Gaussian surface brightness of size 0.5\deg.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in equation~\ref{gauss_pdf} and
equation~\ref{disk_pdf} respectively.} This plot shows that we are
not very sensitive to probing the exact structure of an extended source.
In the search for extended sources, we therefore use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

\subsection{Significance of Extension}
\label{monte_carlo_validation}

\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension as the same conditions
apply (with source flux $\Phi$ replaced by source radius $r$).
To validate this claim, we performed a Monte Carlo
study to calculate empirical distributions for $\tsext$ and
compared them to equation~\ref{ts_ext_distribution}.

We simulated point sources of varying spectral models using
\gtobssim\footnote{\gtobssim is distributed by the Fermi Science Support
Center \citep{fssc}.} and fit them them using \pointlike as both point
and extended sources.  The point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times
10^{-9}$ to $10^{-6} \ph/\cm^2/\sec$ in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameters in LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one year observation period using a default rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1-100 \gev photons and the Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs).  For each point source, we used \pointlike
to fit it as both a point and extended source and calculate \tsext and
we only kept the sources which had a significant point source detection
($\ts>25$).

For most of the spectral models, ~30,000 statistically independent
simulations were performed. For the dimmer spectral models, many of the
simulations left the source undetected ($\ts<25$)
and were discarded.  Table~\ref{ts_ext_num_sims}
shows the different spectral models used in our study as well as the
number of simulations.  The cumulative density of \tsext is plotted in
figure~\ref{ts_ext_mc}. The $\chi^2_1/2$ distribution of
equation~\ref{ts_ext_distribution} is overlaid for comparison.

>>>>

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\renewcommand{\sec}{\text{s}\xspace}
\newcommand{\phflux}{\ensuremath{\ph\ \cm^{-2}\s^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}
\newcommand{\loglikelihood}{\ensuremath{LL}\xspace}
\newcommand{\likelihood}{\ensuremath{L}\xspace}
\newcommand{\chandra}{\text{{\em Chandra}}\xspace}
\newcommand{\swiftxrt}{\text{{\em Swift}/XRT}\xspace}
\newcommand{\rosat}{\text{{\em ROSAT}}\xspace}
\newcommand{\suzaku}{\text{{\em Suzaku}}\xspace}
\newcommand{\asca}{\text{{\em ASCA}}\xspace}
\newcommand{\einstein}{\text{{\em Einstein}}\xspace}
\newcommand{\xmmnewton}{\text{{\em XMM-Newton}}\xspace}

@@

\begin{abstract}
We present a new method for analyzing 
the spatial extension of sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em Fermi Gamma-ray Space Telescope} (FGST). 
Spatial extension
is an important input to correctly 
associate LAT sources with their counterparts at other
wavelengths. We perform
a series of Monte Carlo studies to validate this tool and calculate
the LAT's detection threshold to detect the extension of 
spatially extended sources.  We then
apply this tool to test all sources in the two year source catalog for
extension. We report the detection of nine spatially extended sources
in addition to the twelve spatially extended sources reported in the
second Fermi-LAT Catalog (2FGL).
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the Fermi
Gamma-Ray Space Telescope (FGST) that been surveying the $\gamma$-ray
sky since June 2008.  The LAT has a broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\approx 2.4 \sr$), and large effective area
($\approx 8000 \cm^2$ at $>1 \gev$).

Using two years of all-sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources \citep{second_cat}. Many of these can be
positionally associated with a variety of source classes including Active
Galactic Nuclei (AGN), pulsars, galaxies, and supernova remnants (SNRs).
While the counterparts of many LAT sources
can be spatially resolved when observed
at other frequencies, detection of an extension at \gev energies by
the LAT is complicated by the large point-spread function (PSF)
of the instrument. The angular resolution for events converting in the
thin conversion foils of the tracker system, defined as the 68\% containment radius for
a point source, is approximately $3.5^{\circ}$ at 100 \mev, improving
to about 0.1$^{\circ}$ at 10 \gev.

However, a number of sources positionally coincident with LAT sources
exhibit extensions at other wavelengths which are comparable to or larger than 
the LAT PSF.  In particular, certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites can therefore be expected to be detected as spatially extended
sources by the LAT.  
The LAT collaboration has previously reported five SNRs which are
spatially extended beyond the LAT PSF at \gev energies 
(IC443, W28, W44,
W51C, and RX J1713.7-3946 \citep{ic443,w28,w44,w51c,rx_j1713_lat}). In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 \citep{msh1552,velax,fermi_hess_j1825}, as well
as two close-by galaxies, the Large Magellanic and Small Magalenic
Cloud \citep{lmc,smc}, and finally the radio galaxy Centarus A
\citep{cen_a_lat}.

For each of these sources, the extension was determined based on
dedicated analysis of the particular source 
instead of being the outcome of a systematic search of the LAT catalog.
For the systematic scan of all LAT catalog sources presented in this work, we
therefore expect to find additional spatially extended LAT sources, in
particular since many extended sources in the Galactic plane have been
detected at \tev energies using air Cherenkov detectors and \tev and
\gev emission often originates from the same population of high-energy
particles.  A variety of spatially extended sources have been detected
in the Galactic plane at \tev energies. Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which reported 14 spatially extended sources
 with extensions varying from
$~0.1\deg$ to $~0.25\deg$ \citep{hess_plane_survey}.
In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes}
have 
no significant extension detected at \tev energies. Given
these findings, it can be expected that if the source population at \gev
energies has similar properties, that at least some of the Fermi-LAT
sources in our Galaxy should be spatially resolvable at \gev energies.

@@

The capability to
spatially resolve GeV $\gamma$-ray sources is
important for several reasons. 
First, multiple potential counterparts detected at other frequencies
exist within the localization confidence region of a typical LAT source
along the Galactic plane.
Finding a coherent
source extension across different energy bands allows a firm association
of a LAT source to an otherwise confused counterpart.  Also, due
to the strong energy dependence of the LAT PSF,
the spatial and spectral information
about a source do not simply decouple. An inaccurate spatial model will bias
the spectral model of the source. Specifically, performing a spectral
analysis on a spatially extended source modeled as point-like will
systematically shift the spectral index to be softer.  Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Cloud \citep{first_cat}.
Additionally, such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources in the \gev energy range using the
LAT are challenging because of the strongly energy-dependent PSF.
Additional complications arise for sources along the Galactic plane due to
systematic uncertainties in the Galactic diffuse emission.  The LAT's
PSF is limited at its lowest detectable energies by multiple scattering in the tracker
and the 68\% containment radius of the PSF is approximately 5.1\deg
at 100 \mev  when averaged over instrument acceptance and including
photons which convert in either the thick or thin layers of the
tracker. The PSF improves with energy and approaches a limit given
by the granularity of the silicon strips and is 0.14\deg at 10 \gev
\citep{on_orbit_calibration}.  However, since most high energy
astrophysical sources have power-law spectra, the better resolution of the higher
energy photons is offset by their low
statistics. Therefore a detailed investigation has been performed in this
work to determine the optimal energy range to measure source extension
(section \ref{extended_source_search_method}).

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool
performs a maximum likelihood analysis in which the Poisson likelihood
for observing the measured counts is maximized,
given a parametrized spatial and spectral model of 
the source and its surrounding region.
The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position,
extension, and spectrum can be simultaneously fit.

This analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center \citep{fssc}} because \gtlike can only 
fit the spectral parameters of the model unless a more sophisticated
iterative procedure is used to also test various source morphologies.
We note that \gtlike has been used in the past in several studies of
source extension in the LAT collaboration \citep{lmc,smc,w28,w51c}.
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort has been
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

The choice to base the spatial extension fitting on \pointlike
rather than \gtlike was made on considerations of computing time.
The \pointlike algorithm was optimized for speed to handle
larger numbers of sources efficiently which is important for
our catalog scan as well as the Monte Carlo validation of our method.
Details on the \pointlike package can be found in \cite{matthew_kerr_thesis}.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial
distribution and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source \citep{minuit_documentation}.  As will be described in
section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is important to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated
from \tsext, which 
is the test statistic of the likelihood ratio test of
a model with a spatially extended source and a model
with a point-like source and is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) = 2 (\loglikelihood_\text{ext}-\loglikelihood_\text{ps})
\end{equation}
where \loglikelihood is the log of the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
section~\ref{monte_carlo_validation}.

For extended sources with an assumed radially symmetric shape,
we use a semi-analytic calculation to speed up the calculation of
the expected photon distribution. The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r'^2 dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a {\em King function}
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be evaluated analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r'^2 dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

There will always be a small numerical discrepancy between the expected
photon distribution derived from a true point source and a very small
extended source due to numerical error in the convolution.  In most
situations, this error is insignificant.  But in particular for
very bright sources, this numerical error has the potential to bias the
test statistic for the extension test. Therefore, when calculating
\tsext, we compare the likelihood fitting the source with an extended
spatial model to the likelihood when fixing its extension to ${10^{-10}}\deg$.

We estimate the error on the extension of a source by fixing the position
of the source and varying the extension until the likelihood has fallen
by $\tfrac{1}{2}$, corresponding to a $1\sigma$ error.  This method is
shown schematically in figure~\ref{four_plots_ic443}(a) which shows the
change in \loglikelihood when varying the extension of the SNR IC443.
The localization error is calculated by fixing the extension and spectrum
of the source and fitting to the likelihood function a Gaussian as a function
of position.

@@

\pointlike is important for LAT analyses that require many iterations
such as source localization and extension fitting.  On the other hand,
because of the fewer approximations used in calculating the likelihood
we expect the spectral parameters found with \gtlike to be slightly
more precise.  Furthermore, because \gtlike is the standard likelihood
analysis package, it has been more extensively validated for spectral
analysis.  For those reasons, in the following analysis  we use \pointlike
to determinate the position and extension of a source and subsequently determine
the spectrum using \gtlike. Both \gtlike and \pointlike can be 
used to estimate
the statistical significance of the extension of a source and we require
that both methods agree to claim a detection.

We found good agreement between the two methods.  
Unless explicitly mentioned, all \ts, \tsext, and spectral
parameters were calculated using \gtlike using the best-fit
positions and extensiosn found with \pointlike.

@@

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a two dimensional Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, their convolution
with the PSF look similar.  Figure~\ref{compare_disk_gauss}
demonstrates this by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and the Gaussian surface brightness of size 0.5\deg.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in equation~\ref{gauss_pdf} and
equation~\ref{disk_pdf} respectively.} This plot shows that the LAT is 
not very sensitive to probing the exact structure of an extended source.
In the search for extended sources, we therefore use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

\label{monte_carlo_validation}

We test the performance of \pointlike fitting the extension of
point-like sources.
\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension as the same conditions
apply (with the source flux $\Phi$ replaced by the source radius $r$ and
$r<0$ being unphysical).
To validate this claim, we performed a Monte Carlo
study to calculate empirical distributions for $\tsext$ and
compared them to equation~\ref{ts_ext_distribution}.

We simulated point sources with various spectral forms using
\gtobssim\footnote{\gtobssim is distributed by the Fermi Science Support
Center \citep{fssc}.} and fit them with \pointlike using both point
and extended source hypotheses.  The point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times
10^{-9}$ to $10^{-6}$ \phflux in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameter of LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$
taken to be the same as the isotropic spectrum measured by EGRET
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one year observation period using a default rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1-100 \gev photons and the Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs, \cite{lat_on_orbit_psf}).  For each 
significantly detected point source ($\ts>25$), we used \pointlike
to fit it as an extended source and calculate \tsext.
